Mendelian randomization for causal inference in Genome-wide association studies

### 2022-02-18 mr-review.Rmd
deleted unused R code

@@

Great interest has developed in inferring causal relationships between complex 
traits, *i.e.*, traits that seemingly are not inherited in a Mendelian fashion, in observational human genetics studies. Discovery of such relationships is crucial to 
enhancing our understanding of the biology of health and disease. Two-sample 
Mendelian randomization (MR) studies are a form of instrumental variable analysis in which
genetic markers, typically single nucleotide polymorphisms (SNPs), serve as instruments for 
inferring causal effects of an exposure variable on an outcome variable [@wright1928tariff;@wright1921correlation;@wright1923theory;@stock2003retrospectives;@katan1986apolipoprotein;@katan2004commentary;@didelez2007mendelian]. They are termed 
"two-sample" MR studies because the exposure variable and the outcome variable are measured on 
two non-overlapping cohorts of subjects [@pierce_efficient_2013]. Importantly, the SNP instruments are measured in both sets of subjects.

Genome-wide association studies (GWAS) present unique opportunities for 
discovery of causal relationships via MR. A GWAS
interrogates millions of single nucleotide polymorphisms (SNPs) to infer which affect the trait [@the_wellcome_trust_case_control_consortium_genome-wide_2007;@klein_complement_2005].
In the nearly 17 years since publication of early studies, researchers have reported 
thousands of novel SNP-trait associations from GWAS [@klein_complement_2005;@welter_nhgri_2014;@visscher_10_2017;@visscher_five_2012].

@@

Many MR methods use a set of pre-selected SNPs as instruments. Typically, these SNPs are selected to 
be statistically independent. This restriction to 
independent SNPs is needed for valid inference in methods like inverse variance weighting MR [@burgess_mendelian_2013]. The independent SNPs may be 
selected through linkage disequilibrium (LD) clumping. In this procedure, SNPs 
are ranked by the strength of their marginal association with the exposure 
variable.

LD clumping for choosing instrument SNPs can be problematic for several 
reasons. First, the chosen SNPs may not be causal for the outcome. This can 
happen, for instance, by using a noncausal SNP that is in high LD with a 
causal SNP. Second, because many traits are polygenic or omnigenic, multiple 
causal SNPs may be in high LD together. LD clumping is likely to omit one or more
of the causal SNPs, which can diminish statistical power in MR analysis.

Horizontal pleiotropy occurs when a SNP instrument affects the outcome through at least one pathway that bypasses the exposure variable [@verbanck_detection_2018]. It constitutes a violation of 
the standard MR assumptions and can lead to biased 
causal effect estimates and diminished statistical power. We discuss approaches to 
modeling horizontal pleiotropy below.

```{r timeline, fig.cap="Timeline of two-sample GWAS summary statistics MR methods"}
readxl::read_xlsx("data/method-year.xlsx", sheet = "inputs")
```

@@

The third assumption ensures that the genetic marker is
independent of all confounders of the exposure-outcome relationship.

When the three MR assumptions hold, one may use instrumental variable statistical methods, 
with the genetic marker data, the exposure data, and the outcome data, to estimate and to test the 
causal effect of the exposure on the outcome [@didelez_mendelian_2007;@lawlor_mendelian_2008;@palmer_adjusting_2008]. It's important 
to recognize that assumptions 2 and 3, which require absence of two types of confounding, 
can't be verified in observational studies. For this reason, researchers have developed 
sensitivity analysis to consider the robustness of their inferences to possible violations 
of assumptions 2 and 3.

>>>>

# Abstract

Complex trait genetics studies in humans have yielded thousands of new associations for  
diverse traits. The next step in understanding biology of health and disease is to
identify causal relationships among complex traits. Two-sample Mendelian randomization studies 
in genome-wide association studies can uncover causal relationships 
between complex traits. By leveraging shared SNP genotypes as instrumental variables, or proxies, for 
the exposure complex trait, investigators can tease out causal effects from observational data, provided
that three assumptions are satisfied. We discuss below the growth of genome-wide association studies 
in parallel with development of Mendelian randomization methods. We argue that the recent availability of GWAS summary statistics for diverse complex traits has led to new Mendelian randomization methods and that this area continues to offer opportunities for additional methods developments.

@@

Great interest has developed in inferring causal relationships between complex 
traits, *i.e.*, traits that seemingly are not inherited in a Mendelian fashion, in observational human genetics studies. Discovery of such relationships is crucial to 
enhancing our understanding of the biology of health and disease. Two-sample 
Mendelian randomization (MR) studies are a form of instrumental variable analysis in which
genetic markers, typically single nucleotide polymorphisms (SNPs), serve as instruments, or proxies, for 
inferring causal effects of an exposure variable on an outcome variable [@wright1928tariff;@wright1921correlation;@wright1923theory;@stock2003retrospectives;@katan1986apolipoprotein;@katan2004commentary;@didelez2007mendelian]. They are termed 
"two-sample" MR studies because the exposure variable and the outcome variable are measured on 
two non-overlapping cohorts of subjects [@pierce_efficient_2013]. Importantly, the SNP instruments are measured in both sets of subjects.

Genome-wide association studies (GWAS) present unique opportunities for 
discovery of causal relationships via MR. A GWAS
interrogates millions of single nucleotide polymorphisms (SNPs) to infer which affect the trait [@the_wellcome_trust_case_control_consortium_genome-wide_2007;@klein_complement_2005].
In the nearly 17 years since publication of early studies, researchers have reported 
thousands of novel SNP-trait associations from GWAS [@klein_complement_2005;@welter_nhgri_2014;@visscher_10_2017;@visscher_five_2012]. Ever larger sample sizes in GWAS have enhanced statistical power to detect associations and refined our understanding of 
human health and disease.

@@

Many MR methods use a set of pre-selected SNPs as instruments. Typically, these SNPs are selected to 
be statistically independent. This restriction to 
independent SNPs is needed for valid inference in methods like inverse variance weighting MR [@burgess_mendelian_2013]. The independent SNPs may be 
selected through linkage disequilibrium (LD) clumping.

However, LD clumping for choosing instrument SNPs can be problematic for several 
reasons. First, the chosen SNPs may not be causal for the outcome. This can 
happen, for instance, by using a noncausal SNP that is in high LD with a 
causal SNP. Second, because many traits are polygenic or omnigenic, multiple 
causal SNPs may be in high LD together. LD clumping is likely to omit one or more
of the causal SNPs, which can diminish statistical power in MR analysis.

Horizontal pleiotropy occurs when a SNP instrument affects the outcome through at least one pathway that bypasses the exposure variable [@verbanck_detection_2018]. It constitutes a violation of 
the standard MR assumptions and can lead to biased 
causal effect estimates and diminished statistical power. We discuss approaches to 
modeling and accounting for horizontal pleiotropy below.

@@

The third assumption ensures that the genetic marker is
independent of all confounders of the exposure-outcome relationship. Like the exclusion restriction
assumption, we can't verify this assumption with observational data, so analysts need to 
use tools like sensitivity analysis to ensure validity of findings.

When the three MR assumptions hold, one may use instrumental variable statistical methods, 
with the genetic marker data, the exposure data, and the outcome data, to estimate and to test the 
causal effect of the exposure on the outcome [@didelez_mendelian_2007;@lawlor_mendelian_2008;@palmer_adjusting_2008].
